---
title: "Crime and What Else?"
subtitle: "Investigating the Predictive Power of Social Factors on Incarceration Rates"
author:
  - "Emery Kerani"
date: today
bibliography: refs.bib
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

<!--
  DATA 510 write-up skeleton (Week 5 Quarto lecture).

  Copy this file to your DS3 repo:
    deliverables/M4-writeup-draft/writeup.qmd

  Week 5: paste prose from your M1 proposal into Introduction, Data, Ethics, Methods.
  Week 7+: replace stubs with real results from notebooks/ and src/.
  Week 12 (M4): methods and findings should be stable; render PDF + HTML for submission.
-->

# Title, Team, and Affiliations

**Project title:** *Crime and What Else? Investigating the Predictive Power of Social Factors on Incarceration Rates*

| Name | Role | Willamette email |
|------|------|------------------|
| Emery Kerani | Owner Product Lead | eskerani@willamette.edu |

**Peer Stakeholder POs:** Mary Rose Krouse, Shanti Brodnick, Rohan Srinivas Babu

**Repository:** `https://github.com/eskerani/prison-predictor-capstone/tree/main`

# Abstract

*(150 to 250 words after you have results. Summarize: problem, approach, key finding, impact, limitation. Write this section last.)*

# Introduction: Problem and Research Questions

<!-- Paste and revise from M1 proposal Section 2. Note any changes since M1. -->

**Problem statement:** The United States prison system currently holds nearly 2 million people, and for decades has imprisoned its inhabitants at one of the highest rates in the world (Lurigio, 2024). A variety of legal factors, including mandatory minimum sentences and “three strikes” laws, have contributed to the unprecedented rise in incarceration in the US since the 1970s and 80s (DeFina and Hannon, 2013; Lurigio, 2024). The US incarceration rate has since levelled off, remaining relatively steady since the 2010s; however, this does not reflect the nation’s crime rate, which has fallen since peaking in in the 1990s (Pettit and Gutierrez, 2013). Given that the US’s criminal legal system wields tremendous power—power that is most often turned against already disadvantaged individuals and their communities—it is important that we recognize protective and risk factors for incarceration, many of which are social in nature (Lurigio, 2024; Henry, 2020; Pettit and Gutierrez, 2013). This project seeks to interrogate the dominant narrative that crime is the primary cause of incarceration by examining three other risk factors. 


**Primary research question:** How do the predictive power of social factors (such as poverty, race, or government spending on police) compare to that of violent and property crime rates when using support vector regression to model the national incarceration rate?

**Secondary questions (if any):** If crime is in fact the best predictor of incarceration rates, which social factor is the best predictor of crime rates?

**Changes since M1 proposal:** None yet

# Impact and Stakeholders

The key stakeholders in this work are those who are most often targeted by criminalization and incarceration, primarily poor and BIPOC folks (Pettit and Gutierrez, 2013). Additionally, communities across the country are continually harmed by increasing incarceration rates, as incarceration has been shown to negatively impact both community health (Wildeman and Wang, 2017) and poverty (DeFina and Hannon, 2013). By critically examining the circumstances that can contribute to incarceration, this project seeks to uplift the work of organizations such as Transformative Justice Community and the Oregon Justice Resource Center, which advocate for solutions to harm that shift our paradigm away from imprisonment.

My shared theme is social justice and identity-driven disparity analysis. As others who are interested in social justice but perhaps not as familiar with the prison-industrial complex as I am, I hope that they’ll point out anywhere I make assumptions/lack context and help ensure that my project doesn’t steer anyone toward misinformed conclusions. 

# Related Work and Background

<!-- Brief cited context. Example: -->

Wildfire smoke and respiratory health are well studied at regional scale [@reid2016; @liu2016]. This project applies county-level public data to Oregon specifically.

# Data and Data Engineering

<!-- Paste from M1 Sections 4 and 5; update with actual pipeline state after M2. -->

**Sources:**

| Source | Role | Access |
|--------|------|--------|
| | | |

**Engineering summary:** Raw data in `data/raw/`; processed panel in `data/processed/`; ingest scripts in `src/ingest/`.

```{r}
#| label: tbl-data-qc
#| tbl-cap: "Example QC summary (replace with your panel)"
#| echo: true

# Replace with your project data when ready:
# panel <- readRDS(here::here("data/processed/analysis_panel.rds"))
# nrow(panel)
knitr::kable(data.frame(
  check = c("Rows", "Counties", "Date range"),
  status = c("stub", "stub", "stub")
))
```

# Ethics and Responsible Use

<!-- Paste from M1 Section 8; expand with project-specific limits. -->

- **Privacy:**
- **Fairness / equity:**
- **Limitations of setting:**
- **Non-causal framing:** *(if observational)*

# Methods and Evaluation

<!-- Paste from M1 Section 6; tighten once analysis is fixed. -->

**Design:**

**Primary analysis:**

**Evaluation metrics:**

**Train / validation / holdout:**

```{r}
#| label: fig-methods-placeholder
#| fig-cap: "Replace with a methods diagram or exploratory plot"
#| echo: false
#| fig-width: 5
#| fig-height: 3

plot(1, 1, type = "n", xlab = "", ylab = "", main = "Methods figure placeholder")
text(1, 1, "Replace after M2/M3")
```

# Results and Visualizations

<!-- Fill after M2 data summary and M3 poster figures. -->

*(Stub: key findings with @fig-methods-placeholder and tables.)*

# Discussion, Limitations, and Future Work

**Interpretation:**

**Limitations:** *(tie to actual data and design)*

**Future work:**

# Conclusions and Recommendations

*(Short, stakeholder-facing takeaways.)*

# Reproducibility and References

**Reproduce main results:**

```bash
cd your-repo
quarto render deliverables/M4-writeup-draft/writeup.qmd
```

**Dependencies:** `requirements.txt` or `renv.lock` in repo root.

**Data:** URLs and approval status documented in `docs/data-dictionary.md`.

# References

::: {#refs}
:::

<!-- Create refs.bib alongside this file. Example entries: -->

<!--
@article{liu2016,
  title={Wildfire-specific fine particulate matter and risk of hospital admissions in urban and rural counties},
  author={Liu, J. C. and others},
  journal={Epidemiology},
  year={2016}
}

@article{reid2016,
  title={Critical review of health impacts of wildfire smoke exposure},
  author={Reid, C. E. and others},
  journal={Environmental Health Perspectives},
  year={2016}
}
-->
